---
title: "Toronto’s Housing Costs Stand Out — But It’s Not the Only City Under Pressure(2016–2021"
author: "Ziyan(Bollar) Zhao & Shencen Cai"
output: html_document
---

```{r setup, include=FALSE}
 # install.packages("survey")
 #install.packages("ggrepel")
 #install.packages("tidyverse")
# install.packages("sf")
# install.packages("geojsonsf")


library(geojsonsf)
library(jsonlite)
library(data.table)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyverse)
library(survey)
library(plotly)
library(ggrepel)
library(sf)
library(scales)
library(stringr)
library(htmlwidgets)
library(htmltools)
library(crosstalk)


wmean <- function(x, w) {
  ok <- is.finite(x) & !is.na(x) & is.finite(w) & !is.na(w) & w > 0
  sum(x[ok] * w[ok]) / sum(w[ok])
}

wquantile <- function(x, w, probs = c(0.1, 0.5, 0.9)) {
  ok <- is.finite(x) & is.finite(w) & !is.na(x) & !is.na(w) & w > 0
  x <- x[ok]; w <- w[ok]
  if (length(x) == 0) return(rep(NA_real_, length(probs)))
  o <- order(x)
  x <- x[o]; w <- w[o]
  cw <- cumsum(w) / sum(w)
  sapply(probs, function(p) x[which(cw >= p)[1]])
}
```

# Data and Methods


We use the confidential microdata files from the 2016 and 2021 Censuses, restricted to individuals in Census Metropolitan Areas(CMAs). The key variables are:
CMA: Metropolitan area code (mapped to city names)

AGEGRP: Age group (we select groups 6 and 7, corresponding to ages 25–34)

TotInc: Total income (annual, in dollars)

SHELCO: Total shelter costs (annual, in dollars)

HCORENEED_IND: Core housing need indicator (used for quality checks)

WEIGHT: Survey weight to produce population estimates

Special codes(e.g., 9999999 for income, 888 for core need) are recorded as missing. All estimates are weighted using survey weights, and we report 95% confidence intervals for proportions using the Wilson score methods adapted for survey data. 

```{r cars}
need_cols <- c("CMA", "AGEGRP", "HDGREE", "TotInc", "SHELCO", "HCORENEED_IND", "WEIGHT")

c16 <- fread("census_16_metro.csv", select = need_cols, showProgress = FALSE)
c21 <- fread("census_21_metro.csv", select = need_cols, showProgress = FALSE)

c16$Year <- 2016
c21$Year <- 2021

census <- bind_rows(c16, c21)
```


```{r}
young <- census %>%
  filter(AGEGRP %in% c(9, 10)) %>%
  mutate(
    TotInc_clean = case_when(
      TotInc < 0 ~ NA_real_,
      TotInc >= 9999999 ~ NA_real_,
      TRUE ~ as.numeric(TotInc)
    ),
    SHELCO_clean = case_when(
      SHELCO < 0 ~ NA_real_,
      TRUE ~ as.numeric(SHELCO)
    ),
    HCN_clean = case_when(
      HCORENEED_IND == 888 ~ NA_real_,
      TRUE ~ as.numeric(HCORENEED_IND)
    ),
    AnnualShelter = SHELCO_clean * 12,
    Burden = AnnualShelter / TotInc_clean,
    Burden30 = ifelse(is.finite(Burden) & TotInc_clean > 10000 & Burden > 0.30, 1, 0),
    BachelorPlus = ifelse(HDGREE >= 9 & HDGREE <= 13,1,0),
    City = case_when(
      CMA == 205 ~ "Halifax",
      CMA == 399 ~ "Kingston",
      CMA == 421 ~ "Greater Sudbury",
      CMA == 462 ~ "Montreal",
      CMA == 499 ~ "Sherbrooke",
      CMA == 505 ~ "Ottawa-Gatineau",
      CMA == 532 ~ "Oshawa",
      CMA == 535 ~ "Toronto",
      CMA == 537 ~ "Hamilton",
      CMA == 539 ~ "St. Catharines-Niagara",
      CMA == 541 ~ "Kitchener-Cambridge-Waterloo",
      CMA == 555 ~ "London",
      CMA == 559 ~ "Windsor",
      CMA == 577 ~ "Barrie",
      CMA == 588 ~ "Brantford",
      CMA == 599 ~ "Guelph",
      CMA == 602 ~ "Winnipeg",
      CMA == 799 ~ "Regina",
      CMA == 825 ~ "Calgary",
      CMA == 835 ~ "Edmonton",
      CMA == 933 ~ "Vancouver",
      CMA == 935 ~ "Victoria",
      CMA == 988 ~ "Kelowna",
      TRUE ~ "Other"
    )
  ) %>%
  filter(
    City != "Other",
    WEIGHT > 0,
    !is.na(TotInc_clean),
    !is.na(SHELCO_clean)
  )
```

```{r}
metro_clean <- young %>%
  mutate(
    TotInc = ifelse(TotInc >= 9999999, NA, TotInc),
    HCORENEED_IND = ifelse(HCORENEED_IND == 888, NA, HCORENEED_IND)
  ) %>%
  filter(
    AGEGRP %in% c(9,10),
    TotInc >0, SHELCO >= 0
  )

# to prevent extreme high incomes from affecting ratios, trim the top 1%
metro_clean <- metro_clean %>%
  filter(TotInc < quantile(TotInc, 0.99, na.rm = TRUE))

summary(metro_clean)
```

```{r}
young %>%
  summarise(
    n = n(),
    miss_income = sum(is.na(TotInc_clean)),
    miss_shelter = sum(is.na(SHELCO_clean)),
    miss_hcn = sum(is.na(HCN_clean)),
    miss_weight = sum(is.na(WEIGHT))
  )
```


```{r summaries, message=FALSE, warning=FALSE}
young %>%
  summarise(
    mean_income = weighted.mean(TotInc_clean, WEIGHT, na.rm = TRUE),
    median_income = median(TotInc_clean, na.rm = TRUE),
    mean_shelter = weighted.mean(SHELCO_clean, WEIGHT, na.rm = TRUE),
    burden_rate = weighted.mean(Burden30, WEIGHT, na.rm = TRUE)
  )
```

```{r}
burden_city <- young %>%
  group_by(City, Year) %>%
  summarise(
    burden_rate = wmean(Burden30, WEIGHT),
    mean_income = wmean(TotInc_clean, WEIGHT),
    mean_monthly_shelter = wmean(SHELCO_clean, WEIGHT),
    mean_annual_shelter = wmean(AnnualShelter, WEIGHT),
    median_burden = wquantile(Burden, WEIGHT, 0.5)[1],
    mean_shelter = weighted.mean(SHELCO_clean, WEIGHT, na.rm = TRUE),
    p90_burden = wquantile(Burden, WEIGHT, 0.9)[1],
    pop_weight = sum(WEIGHT, na.rm = TRUE),
    .groups = "drop"
  )

burden_year <- young %>%
  group_by(Year) %>%
  summarise(
    burden_rate = weighted.mean(Burden30, WEIGHT, na.rm = TRUE),
    mean_income = weighted.mean(TotInc_clean, WEIGHT, na.rm = TRUE),
    mean_shelter = weighted.mean(SHELCO_clean, WEIGHT, na.rm = TRUE)
  )
```


```{r}
ggplot(young %>% filter(Burden >= 0, Burden < 5), aes(Burden)) +
  geom_histogram(bins = 60, fill = "steelblue") +
  geom_vline(xintercept = 0.30, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Housing Cost Burden (Age 25–34)",
    x = "Shelter Cost / Income",
    y = "Count"
  )

ggplot(young %>% filter(Burden >=0, Burden <5),
       aes(x = Burden, fill = factor(Year))) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0.30, linetype = "dashed") +
  labs(
    title = "Distribution of Housing Burden Over Time",
    x = "Shelter Cost / Income",
    fill = "Year"
  )
```


```{r}
ggplot(burden_city,
       aes(x = reorder(City, burden_rate), y = burden_rate, fill = factor(Year))) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Housing Cost Burden Among Young Adults by City",
    x = NULL,
    y = "Share spending >30% of income",
    fill = "Year"
  )
```
```{r}
ggplot(burden_city, aes(x = mean_income, y = burden_rate, color = factor(Year))) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = City), size = 3) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Housing Cost Burden vs. Mean Income by City",
    x = "Mean Total Income (CAD)",
    y = "Share spending >30% of income",
    color = "Year"
  )
```
```{r}
rent_city <- young %>%
  group_by(City, Year) %>%
  summarise(
    mean_shelter = weighted.mean(SHELCO_clean, WEIGHT)
  )

ggplot(rent_city,
       aes(x = reorder(City, mean_shelter), y = mean_shelter, fill = factor(Year))) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Average Shelter Costs by City",
    x = NULL,
    y = "Monthly Shelter Cost"
  )
```

# Toronto's housing costs stand out--but it's not the only city under pressure

# Opening
reality meaning of our topic + briefly introduction of the data cleaning process, data information(source, key variables, etc.)......

# Lead out to first question and visualization 1(map): Where do young adults face the highest housing pressure?

```{r}

#统一city name
cma_sf <- st_read("lcma000b16a_e/lcma000b16a_e.shp", quiet = TRUE) %>%
  st_transform(4326) %>%
  select(CMANAME, geometry)

clean_name <- function(x) {
  x %>%
    str_trim() %>%
    str_squish()
}

cma_sf2 <- cma_sf %>%
  mutate(CMANAME = clean_name(CMANAME))

burden_city2 <- burden_city %>%
  mutate(
    CMANAME = clean_name(City),
    CMANAME = recode(
      CMANAME,
      "Greater Sudbury" = "Greater Sudbury / Grand Sudbury",
      "Kitchener-Cambridge-Waterloo" = "Kitchener - Cambridge - Waterloo",
      "Montreal" = "Montréal",
      "Ottawa-Gatineau" = "Ottawa - Gatineau (Ontario part / partie de l'Ontario)",
      "St. Catharines-Niagara" = "St. Catharines - Niagara"
    )
  )

sf_2016 <- cma_sf2 %>%
  left_join(burden_city2 %>% filter(Year == 2016), by = "CMANAME") %>%
  filter(!is.na(burden_rate))

sf_2021 <- cma_sf2 %>%
  left_join(burden_city2 %>% filter(Year == 2021), by = "CMANAME") %>%
  filter(!is.na(burden_rate))

sf_2016_fixed <- sf_2016 %>%
  st_make_valid()

sf_2016_s <- sf_2016_fixed %>%
  st_simplify(dTolerance = 1000)

sf_2021_fixed <- sf_2021 %>%
  st_make_valid()

sf_2021_s <- sf_2021_fixed %>%
  st_simplify(dTolerance = 1000)

tmp2016 <- tempfile(fileext = ".geojson")
tmp2021 <- tempfile(fileext = ".geojson")

sf::st_write(sf_2016_s, tmp2016, delete_dsn = TRUE, quiet = TRUE)
sf::st_write(sf_2021_s, tmp2021, delete_dsn = TRUE, quiet = TRUE)

geojson_2016 <- jsonlite::fromJSON(tmp2016, simplifyVector = FALSE)
geojson_2021 <- jsonlite::fromJSON(tmp2021, simplifyVector = FALSE)
```


```{r}
map_2016 <- sf_2016_s %>%
  st_drop_geometry() %>%
  mutate(
    hover_text = paste0(
      "<b>", CMANAME, "</b>",
      "<br>Year: 2016",
      "<br>Share facing housing pressure: ", percent(burden_rate, accuracy = 0.1),
      "<br>Typical annual income: ", dollar(mean_income),
      "<br>Typical monthly housing cost: ", dollar(mean_shelter)
    )
  )

map_2021 <- sf_2021_s %>%
  st_drop_geometry() %>%
  mutate(
    hover_text = paste0(
      "<b>", CMANAME, "</b>",
      "<br>Year: 2021",
      "<br>Share facing housing pressure: ", percent(burden_rate, accuracy = 0.1),
      "<br>Typical annual income: ", dollar(mean_income),
      "<br>Typical monthly housing cost: ", dollar(mean_shelter)
    )
  )

fig_map <- plot_ly() %>%
  
  add_trace(
    type = "choroplethmapbox",
    geojson = geojson_2016,
    locations = map_2016$CMANAME,
    z = map_2016$burden_rate,
    featureidkey = "properties.CMANAME",
    text = map_2016$hover_text,
    hoverinfo = "text",
    colorscale = "YlOrRd_r",
    zmin = min(c(map_2016$burden_rate, map_2021$burden_rate), na.rm = TRUE),
    zmax = max(c(map_2016$burden_rate, map_2021$burden_rate), na.rm = TRUE),
    marker = list(line = list(color = "white", width = 0.6)),
    visible = TRUE,
    showscale = TRUE,
    colorbar = list(
      title = "Housing pressure",
      tickformat = ".0%"
    )
  ) %>%
  
  add_trace(
    type = "choroplethmapbox",
    geojson = geojson_2021,
    locations = map_2021$CMANAME,
    z = map_2021$burden_rate,
    featureidkey = "properties.CMANAME",
    text = map_2021$hover_text,
    hoverinfo = "text",
    colorscale = "YlOrRd_r",
    zmin = min(c(map_2016$burden_rate, map_2021$burden_rate), na.rm = TRUE),
    zmax = max(c(map_2016$burden_rate, map_2021$burden_rate), na.rm = TRUE),
    marker = list(line = list(color = "white", width = 0.6)),
    visible = FALSE,
    showscale = FALSE
  ) %>%
  
  layout(
    title = list(
      text = "Housing pressure among young adults across Canadian metropolitan areas",
      x = 0.02
    ),
    mapbox = list(
      style = "carto-positron",
      zoom = 2.7,
      center = list(lon = -96, lat = 56)
    ),
    margin = list(l = 0, r = 0, t = 60, b = 0),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0.02,
        y = 1.05,
        buttons = list(
          list(
            method = "update",
            args = list(
              list(visible = list(TRUE, FALSE)),
              list(title = list(text = "Housing pressure among young adults across Canadian metropolitan areas (2016)"))
            ),
            label = "2016"
          ),
          list(
            method = "update",
            args = list(
              list(visible = list(FALSE, TRUE)),
              list(title = list(text = "Housing pressure among young adults across Canadian metropolitan areas (2021)"))
            ),
            label = "2021"
          )
        )
      )
    ),
    annotations = list(
      list(
        x = 0.02, y = 0.98,
        xref = "paper", yref = "paper",
        text = "Darker areas indicate a larger share of young adults spending more than 30% of income on housing.",
        showarrow = FALSE,
        align = "left",
        font = list(size = 12, color = "gray30")
      )
    )
  )

toronto_2021 <- sf_2021_s %>%
  filter(CMANAME == "Toronto") %>%
  st_centroid()

toronto_coords <- st_coordinates(toronto_2021)

fig_map <- fig_map %>%
  add_trace(
    type = "scattermapbox",
    mode = "text",
    lon = toronto_coords[,1],
    lat = toronto_coords[,2],
    text = "Toronto",
    textfont = list(color = "#B22222", size = 14),
    showlegend = FALSE,
    inherit = FALSE
  )

updatemenus = list(
  list(
    type = "buttons",
    direction = "right",
    x = 0.02,
    y = 1.05,
    buttons = list(
      list(
        method = "update",
        args = list(
          list(visible = list(TRUE, FALSE, TRUE)),
          list(title = list(text = "Housing pressure among young adults across Canadian metropolitan areas (2016)"))
        ),
        label = "2016"
      ),
      list(
        method = "update",
        args = list(
          list(visible = list(FALSE, TRUE, TRUE)),
          list(title = list(text = "Housing pressure among young adults across Canadian metropolitan areas (2021)"))
        ),
        label = "2021"
      )
    )
  )
)

fig_map
```

# Key takeaways

# After seeing where the problem is worst, reader ask： Has this always been the case, and which place's housing pressure rose the most?(Lead to visualization 2)


```{r}
#Visualization 2
change_city <- burden_city %>%
  select(City, Year, burden_rate) %>%
  pivot_wider(
    names_from = Year,
    values_from = burden_rate,
    names_prefix = "y"
  ) %>%
  mutate(
    change_pp = 100 * (y2021 - y2016)
  ) %>%
  arrange(y2021) %>%
  mutate(City = factor(City, levels = City))

fig2 <- plot_ly()

fig2 <- fig2 %>%
  add_segments(
    data = change_city,
    x = ~y2016,
    xend = ~y2021,
    y = ~City,
    yend = ~City,
    line = list(color = "rgba(120,120,120,0.45)", width = 2),
    hoverinfo = "none",
    showlegend = FALSE
  ) %>%
  add_markers(
    data = change_city,
    x = ~y2016,
    y = ~City,
    marker = list(
      size = 9,
      color = ifelse(change_city$City == "Toronto", "#D84B4B", "#F28E8B")
    ),
    customdata = ~paste0(
      "City: ", City,
      "<br>2016 share facing housing pressure: ", percent(y2016, accuracy = 0.1),
      "<br>2021 share facing housing pressure: ", percent(y2021, accuracy = 0.1),
      "<br>Change: ", sprintf("%+.1f", change_pp), " percentage points"
    ),
    hovertemplate = "%{customdata}<extra></extra>",
    name = "2016"
  ) %>%
  add_markers(
    data = change_city,
    x = ~y2021,
    y = ~City,
    marker = list(
      size = 9,
      color = ifelse(change_city$City == "Toronto", "#D84B4B", "#18B7BE")
    ),
    customdata = ~paste0(
      "City: ", City,
      "<br>2016 share facing housing pressure: ", percent(y2016, accuracy = 0.1),
      "<br>2021 share facing housing pressure: ", percent(y2021, accuracy = 0.1),
      "<br>Change: ", sprintf("%+.1f", change_pp), " percentage points"
    ),
    hovertemplate = "%{customdata}<extra></extra>",
    name = "2021"
  ) %>%
  layout(
    title = "Housing pressure rose in most cities between 2016 and 2021",
    xaxis = list(
      title = "Share spending more than 30% of income on housing",
      tickformat = ".0%"
    ),
    yaxis = list(title = ""),
    legend = list(orientation = "h", x = 0.35, y = 1.08)
  )

fig2
```

# Takeaways:

# Once reader see that housing pressure increase in many places, they will ask: Why did housing pressure increase?

# Lead to Visualization 3: Why did housing pressure increase overall?

```{r}
# Visualization 3
growth_city <- burden_city %>%
  select(City, Year, mean_income, mean_shelter) %>%
  pivot_wider(
    names_from = Year,
    values_from = c(mean_income, mean_shelter)
  ) %>%
  mutate(
    income_growth = (mean_income_2021 - mean_income_2016) / mean_income_2016,
    shelter_growth = (mean_shelter_2021 - mean_shelter_2016) / mean_shelter_2016,
    highlight = ifelse(City == "Toronto", "Toronto", "Other"),
    label = ifelse(City == "Toronto", "Toronto", "")
  )

max_val <- max(c(growth_city$income_growth, growth_city$shelter_growth))

fig_growth <- plot_ly(
  growth_city,
  x = ~income_growth,
  y = ~shelter_growth,
  type = "scatter",
  mode = "markers+text",
  text = ~label,
  textposition = "top center",
  color = ~highlight,
  colors = c("Other" = "#8ecae6", "Toronto" = "#d84b4b"),
  marker = list(size = 10),
  hovertext = ~paste0(
    "<b>", City, "</b>",
    "<br>Income growth: ", percent(income_growth, accuracy = 0.1),
    "<br>Housing cost growth: ", percent(shelter_growth, accuracy = 0.1)
  ),
  hoverinfo = "text"
) %>%
layout(
  title = "In most cities, housing costs grew faster than incomes",
  
  xaxis = list(
    title = "Income growth (2016–2021)",
    tickformat = ".0%"
  ),
  
  yaxis = list(
    title = "Housing cost growth (2016–2021)",
    tickformat = ".0%"
  ),
  
  shapes = list(
    
    # equality line
    list(
      type = "line",
      x0 = 0, y0 = 0,
      x1 = max_val, y1 = max_val,
      line = list(
        dash = "dash",
        color = "gray50",
        width = 2
      )
    )
    
  ),
  
  annotations = list(
    
    list(
      x = max_val*0.75,
      y = max_val*0.85,
      text = "Above the line:<br>Housing costs grew faster",
      showarrow = FALSE,
      font = list(size = 12, color = "gray30")
    ),
    
    list(
      x = max_val*0.75,
      y = max_val*0.55,
      text = "Below the line:<br>Income grew faster",
      showarrow = FALSE,
      font = list(size = 12, color = "gray30")
    )
    
  )
)

fig_growth
```

# Interpretation
Toronto’s housing pressure may remain severe, but its recent growth pattern is not the clearest example of housing costs outrunning income.

# City-level growth patterns do not fully explain who is under pressure. Even in places where income growth kept pace with or exceeded housing cost growth, many young adults may still be clustered near the affordability threshold. That is why the next chart looks at the distribution of housing burden among individuals.

```{r}
# Visualization 4
dist_city <- young %>%
  filter(Burden >= 0, Burden < 5)

# 2. Weighted density helper
weighted_density_df <- function(data, city_name, year_value, n = 512) {
  d <- data %>%
    filter(City == city_name, Year == year_value)

  dens <- density(
    x = d$Burden,
    weights = d$WEIGHT / sum(d$WEIGHT, na.rm = TRUE),
    from = 0,
    to = 5,
    n = n
  )

  tibble(
    City = city_name,
    Year = year_value,
    burden = dens$x,
    density = dens$y
  )
}

# 3. Build density data for all cities
city_list <- sort(unique(dist_city$City))

density_all <- bind_rows(
  lapply(city_list, function(ct) weighted_density_df(dist_city, ct, 2016)),
  lapply(city_list, function(ct) weighted_density_df(dist_city, ct, 2021))
)

# 4. Choose default city
default_city <- "Toronto"

# 5. Create interactive plot
fig_dist <- plot_ly()

for (ct in city_list) {
  
  d2016 <- density_all %>% filter(City == ct, Year == 2016)
  d2021 <- density_all %>% filter(City == ct, Year == 2021)
  
  fig_dist <- fig_dist %>%
    add_lines(
      data = d2016,
      x = ~burden,
      y = ~density,
      name = "2016",
      visible = (ct == default_city),
      line = list(color = "#F4A6A3", width = 3),
      hovertemplate = paste0(
        "<b>", ct, "</b>",
        "<br>Year: 2016",
        "<br>Share of income spent on housing: %{x:.2f}",
        "<br>Density: %{y:.2f}",
        "<extra></extra>"
      ),
      showlegend = (ct == default_city)
    ) %>%
    add_lines(
      data = d2021,
      x = ~burden,
      y = ~density,
      name = "2021",
      visible = (ct == default_city),
      line = list(color = "#18B7BE", width = 3),
      hovertemplate = paste0(
        "<b>", ct, "</b>",
        "<br>Year: 2021",
        "<br>Share of income spent on housing: %{x:.2f}",
        "<br>Density: %{y:.2f}",
        "<extra></extra>"
      ),
      showlegend = (ct == default_city)
    )
}

# 6. Visibility logic for dropdown
visibility_list <- lapply(seq_along(city_list), function(i) {
  v <- rep(FALSE, length(city_list) * 2)
  v[(2*i - 1):(2*i)] <- TRUE
  v
})

buttons <- lapply(seq_along(city_list), function(i) {
  list(
    method = "update",
    args = list(
      list(visible = visibility_list[[i]]),
      list(
        title = paste0(
          "How housing pressure is distributed among young adults in ", city_list[i]
        )
      )
    ),
    label = city_list[i]
  )
})

# 7. Final layout
fig_dist <- fig_dist %>%
  layout(
    title = paste0(
      "How housing pressure is distributed among young adults in ", default_city
    ),
    xaxis = list(
      title = "Share of income spent on housing",
      range = c(0, 5)
    ),
    yaxis = list(
      title = "Density"
    ),
    updatemenus = list(
      list(
        type = "dropdown",
        x = 1.02,
        y = 1,
        xanchor = "left",
        buttons = buttons
      )
    ),
    shapes = list(
      list(
        type = "line",
        x0 = 0.30, x1 = 0.30,
        y0 = 0, y1 = 1,
        yref = "paper",
        line = list(color = "red", dash = "dash", width = 2)
      )
    ),
    annotations = list(
      list(
        x = 0.30,
        y = 0.95,
        yref = "paper",
        text = "30% affordability line",
        showarrow = FALSE,
        font = list(color = "red", size = 11)
      ),
      list(
        x = 0.98,
        y = 0.92,
        xref = "paper",
        yref = "paper",
        text = "Select a city",
        showarrow = FALSE,
        font = list(size = 12, color = "gray35")
      )
    ),
    legend = list(
      orientation = "h",
      x = 0.35,
      y = 1.08
    )
  )

fig_dist
```
# But city averages can hide important differences. Readers may now ask: Which groups are more protected or exposed?

```{r}
# Visualization 5
edu_gap_city <- young %>%
  mutate(
    Education = ifelse(
      BachelorPlus == 1,
      "Bachelor's degree or higher",
      "No bachelor's degree"
    )
  ) %>%
  group_by(City, Year, Education) %>%
  summarise(
    burden_rate = weighted.mean(Burden30, WEIGHT, na.rm = TRUE),
    mean_shelter = weighted.mean(SHELCO_clean, WEIGHT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(City, Year, Education, burden_rate) %>%
  pivot_wider(
    names_from = Education,
    values_from = burden_rate
  ) %>%
  rename(
    burden_bachelor = `Bachelor's degree or higher`,
    burden_no_bachelor = `No bachelor's degree`
  ) %>%
  mutate(
    education_gap = burden_no_bachelor - burden_bachelor
  ) %>%
  left_join(
    burden_city %>% select(City, Year, burden_rate, mean_shelter),
    by = c("City", "Year")
  ) %>%
  mutate(
    highlight = ifelse(City == "Toronto", "Toronto", "Other"),
    label = ifelse(City == "Toronto", "Toronto", "")
  )

fig_edu_gap <- plot_ly(
  data = edu_gap_city,
  x = ~burden_rate,
  y = ~education_gap,
  frame = ~Year,
  type = "scatter",
  mode = "markers+text",
  text = ~label,
  textposition = "top center",
  color = ~highlight,
  colors = c("Other" = "#8ecae6", "Toronto" = "#d84b4b"),
  size = ~mean_shelter,
  sizes = c(10, 30),
  hovertext = ~paste0(
    "<b>", City, "</b>",
    "<br>Overall housing pressure: ", percent(burden_rate, accuracy = 0.1),
    "<br>Education gap: ", percent(education_gap, accuracy = 0.1),
    "<br>Bachelor's degree or higher: ", percent(burden_bachelor, accuracy = 0.1),
    "<br>No bachelor's degree: ", percent(burden_no_bachelor, accuracy = 0.1),
    "<br>Typical monthly housing cost: ", dollar(mean_shelter)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "In some cities, education offers more protection than in others",
    xaxis = list(
      title = "Overall share facing housing pressure",
      tickformat = ".0%"
    ),
    yaxis = list(
      title = "Gap between no bachelor's degree and bachelor's degree",
      tickformat = ".0%"
    )
  ) %>%
  animation_opts(1000, easing = "cubic-in-out") %>%
  animation_slider(currentvalue = list(prefix = "Year: "))

fig_edu_gap
```

```{r}
# shared city summary
shared_city <- SharedData$new(
  burden_city %>% mutate(City = as.character(City)),
  key = ~City,
  group = "cities"
)

# chart A: burden rate by year
p_a <- plot_ly(
  shared_city,
  x = ~factor(Year),
  y = ~burden_rate,
  color = ~City,
  type = "bar",
  colors = "Set2",
  hovertemplate = paste0(
    "<b>%{customdata}</b>",
    "<br>Year: %{x}",
    "<br>Burden rate: %{y:.1%}",
    "<extra></extra>"
  ),
  customdata = ~City,
  showlegend = FALSE
) %>%
  layout(
    title = "Housing burden by year",
    xaxis = list(title = ""),
    yaxis = list(title = "Burden rate", tickformat = ".0%")
  )

# chart B: income vs shelter
p_b <- plot_ly(
  shared_city,
  x = ~mean_income,
  y = ~mean_shelter,
  type = "scatter",
  mode = "markers+text",
  text = ~ifelse(City == "Toronto", City, ""),
  textposition = "top center",
  marker = list(size = 11, opacity = 0.8),
  hovertemplate = paste0(
    "<b>%{customdata}</b>",
    "<br>Income: %{x:$,.0f}",
    "<br>Monthly shelter: %{y:$,.0f}",
    "<extra></extra>"
  ),
  customdata = ~City
) %>%
  layout(
    title = "Income vs monthly shelter cost",
    xaxis = list(title = "Mean annual income"),
    yaxis = list(title = "Mean monthly shelter cost")
  )

# prepare distribution data by city-year
dist_df <- young %>%
  filter(Burden >= 0, Burden < 5) %>%
  group_by(City, Year) %>%
  mutate(weight_share = WEIGHT / sum(WEIGHT, na.rm = TRUE)) %>%
  ungroup()

shared_dist <- SharedData$new(
  dist_df,
  key = ~City,
  group = "cities"
)

# chart C: burden histogram
p_c <- plot_ly(
  shared_dist,
  x = ~Burden,
  color = ~factor(Year),
  colors = c("#f28e8b", "#18b7be"),
  type = "histogram",
  opacity = 0.6,
  nbinsx = 40
) %>%
  layout(
    title = "Burden distribution",
    barmode = "overlay",
    xaxis = list(title = "Shelter cost / income"),
    yaxis = list(title = "Count"),
    shapes = list(
      list(
        type = "line",
        x0 = 0.30, x1 = 0.30,
        y0 = 0, y1 = 1,
        xref = "x", yref = "paper",
        line = list(color = "red", dash = "dash")
      )
    )
  )

bscols(
  widths = c(3, 9),
  div(
    style = "padding:10px",
    tags$h4("Explore Cities"),
    filter_select(
      id = "city_filter",
      label = "Select a city",
      sharedData = shared_city,
      group = ~City
    )
  ),
  bscols(widths = c(4,4,4), p_a, p_b, p_c)
)
```

#Conclusion & Limitation

